Sequential rank agreement methods for comparison of ranked lists

### 2015-04-21 latex/paper.Rnw
Intro draft

Key words: sequential rank agreement, rank agreement, order statistic,
methods comparison, variable selection

@@

In this article we propose a method to measure agreement among ranked
lists. The proposed method allows for multiple lists simultaneously,
provides a dynamic measure of agreement for each depth in the lists,
places more weight on the top of the list, accommodates censored/incomplete
lists, and has a natural interpretation that directly relates to the
ranks. The general idea is to define agreement based on the ranks of
the first $k$ elements in each list.

The analysis of rankings is hardly new, and several other approaches
have proposed a measure of the ``distance'' between two ranked
lists. Of these, Kendall's $\tau$ \citep{kendall1948} and Spearman’s
$\rho$ \citep{Spearman1910} are the most well-known. Both correlations
measure the similarity of two ranked lists on a correlation scale from
-1 to 1, but neither of them make any distinction between errors
towards the top versus the errors towards the bottom of the lists.
Several modified versions of Kendall's $\tau$ and Spearman's $\rho$
have been proposed. \citet{Shieh1998} proposed a weighted $\tau$ where
each pair of rankings can be assigned different weights, and
\citet{Yilmaz2008} proposed the $\tau_{ap}$ which places higher
emphasis on the top of the lists.
%
Spearman’s $\rho$ use the ranks of the variables for calculation of
the correlation and the use of ranks are also employed in the $M$
measure of \citet{Bar-Ilan2006} where the reciprocal rank differences
are used to calculate the similarity measure.

Recent approaches consider the intersection of the lists as the basis
for a similarity measure

\citetCarterette2009}

Rather than looking at correlations between lists some recent papers
use the intersection of the lists as the basis for a similarity
measure. It can be argued that intersection is a better measure when
looking at non-conjoint lists, but the simple intersection places
equal weights on all depths of the list. Thus, \citet{Fagin2003} and
Webber et al. (2010) attempt to modify the simple intersection in a
way that is top weighted by using the cumulative intersection divided
by the depth as the basis for their similarity measures — in the
following referred to as the average overlap (AO). Webber et
al. (2010) define their rank-biased overlap (RBO) by weighting with a
converging series thus ensuring the top is always weighted higher than
the (potentially indefinite) bottom of the lists. The intersection
metric by Fagin et al. (2003) uses the mean of the overlaps at all
depths up to d.  While it is possible to calculate measures that
describe how similar lists are at a given depth with some of the
methods described above, the interpretation of the measures is unclear
and as we will demonstrate it is not trivial to identify a depth at
which the lists are no longer similar. We will use some terminology
from agreement analysis as our 2 method resembles this in some
aspects. We will refer to the similar part of the lists, which is
expected at the top if it exists, as having high agreement and the
dissimilar part, where items are ranked randomly, as having low
agreement. As a measure for agreement we will base this on the
limits-of-agreement known from agreement between quantitative
variables (Carstensen, 2011) . Since the limit-of-agreement
essentially is a transformation of the pooled standard deviation of
the ranked items we will use that, since it has a direct and simple
interpretation: at given depth how far do the items deviate from the
mean rank. Thus, twice the standard deviation is a measure of how far
apart in ranks the methods classify the items.  The rest of the paper
is structured as follows, in the methods section we show how to
calculate and apply the limits-of-agreement-based measure, sd, to a
small example dataset and we perform a comparison to average
overlap. We also present and discuss how changepoint methods can be
applied to identify where a set of lists goes from high agreement to
low agreement. The simulation section shows the performance of the
measure under a number of different scenarios while the application
shows calculations of agreement and average overlap on a publicly
available colorectal cancer dataset. Finally the findings are
discussed along with possible extensions.

The manuscript is organized as follows: In the next section we will
define ...

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings.

@@

\subsection{All lists fully observed}
We shall start by the simplest case where all $L$ lists are fully
observed, \ie, we have the rank of all $P$ items for all of the $L$
lists. Fully observed ranked lists are common and occur, for example,
when we have a single dataset and apply different statistical analysis
methods to produce lists of predictors ranked according to their
importance, or if the same analysis method is applied to data from
different populations.

@@

\subsection{Analysis of top $k$ lists}
Not uncommon for lists to be

>>>>

Key words: sequential rank agreement, top lists, rank agreement, order statistic,
methods comparison, variable selection

@@

The analysis of rankings is hardly new, and several other approaches
have proposed a measure of the ``distance'' between two ranked
lists. Of these, Kendall's $\tau$ 
%\citep{kendall1948} 
and Spearman’s footrule
%$\rho$ \citep{Spearman1910} 
are the most well-known. Both measure the similarity of two ranked
lists but neither of them make any distinction between errors towards
the top versus the errors towards the bottom of the lists.  Several
modified versions of Kendall's $\tau$ and Spearman's footrule have been
proposed. \citet{Shieh1998} proposed a weighted $\tau$ where each pair
of rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
%
Spearman’s footrule use the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$
measure of \citet{Bar-Ilan2006} where the reciprocal rank differences
are used to calculate the similarity measure.

More recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersections also places
equal weights on all depths of the list which led \citet{Fagin2003}
and \citet{Webber2010} to modify the simple intersection to put more
emphasis on the top of the lists by using the cumulative intersection
divided by the depth as their similarity measures (denoted the average
overlap (AO)). \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series thus ensuring the top is
always weighted higher than the (potentially indefinite) bottom of the
lists.  While it is possible to calculate measures that describe how
similar lists are at a given depth with some of the methods described
above, the interpretation of the measures is unclear --- especially
with multiple lists --- and it is difficult to identify a depth at
which the lists are no longer similar.

In this article we propose a method to measure similarity among ranked
lists for each depth in the list. As a measure for similarity we will
use the limits of agreement known from agreement between quantitative
variables \citep{Carstensen2011} but will compute the agreement on the
ranks of the items.  The proposed method allows for multiple lists
simultaneously, provides a dynamic measure of agreement for each depth
in the lists, places more weight on the top of the list, accommodates
censored/incomplete lists of varying lengths, and has a natural
interpretation that directly relates to the ranks. The general idea is
to define agreement based on the sequence of ranks from the first $k$
elements in each list. This approach allows us to estimate a potential
change point in the agreement in order to determine an optimal $k$
where there is a shift in the agreement among lists and enables us to
use randomization approaches to evaluate the rank agreement. In this
sense it is a combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches.

The manuscript is organized as follows: In the next section we will
define the sequential rank agreement among multiple ranked lists and
discuss how to handle incomplete/censored lists. In section 3 we
present and discuss approaches to evaluate the sequential rank
agreement obtained from an experiment and to identify the optimal
depth $k$ for which the ranked lists agree. Finally we present three
quite different applications of the proposed sequential rank agreement
approach to real data before we discuss the findings along with
possible extensions.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$.

@@

\subsection{Agreement among fully observed lists}
We shall start by the simplest case where all $L$ lists are fully
observed, \ie, we have the rank of all $P$ items for all of the $L$
lists. Fully observed ranked lists are common and occur, for example,
when we have a single dataset and apply different statistical analysis
methods to produce lists of predictors ranked according to their
importance, or if the same analysis method is applied to data from
different populations.

@@

\subsection{Analysis of incomplete/censored lists}
Not uncommon for lists to be
